---
title: "On Singular Value Decomposition"
author: "Miguel Conde"
date: "25 de julio de 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Any $m \times n$ matrix A can be decomposed as:

$$
A_{m \times n} = U_{m \times m} D_{m \times n} V_{n \times n}^T
$$
If we define:
$$
B_{n \times n} = A_{m \times n}^T A_{m \times n}
$$
And:
$$
C_{m \times m} = A_{m \times n} A_{m \times n}^T
$$

Then:

* $D_{m \times n}$ is a diagonal matrix whose diagonal is made of the square roots of the eigenvalues of $B_{n \times n}$, ordered in decreasing order. There can be until $n$ distinct eigenvalues $\lambda_i$, but usually only $r \leq n$ of which will be nonzero.

$$
\lambda_i = \sigma_i^2 \\
\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_r > 0 \\
r \leq n \\
d_1= \sigma_1, d_2 = \sigma_2, ... , d_r = \sigma_r,d_{r+1}= ...= d_{n} = 0, d_{n+1} =... = d_m = 0
$$

* $V_{n \times n}$ is made of the eigenvectors of $B_{n \times n}$

* $U_{m \times m}$ is made of the eigenvectors of $C_{m \times m}$


## Example
```{r}
m <- 4
n <- 3
# A_m_n <- matrix(c(1,2,3,4,5,6,7,8,9, 10, 11, 12), ncol = n)
A_m_n <- matrix(rnorm(m*n), ncol = n)
A_m_n
```

```{r}
B_n_n <- t(A_m_n) %*% A_m_n 
C_m_m <- A_m_n %*% t(A_m_n)
eigen_B_n_n <- eigen(B_n_n, symmetric = isSymmetric(A_m_n))
eigen_C_m_m <- eigen(C_m_m, symmetric = isSymmetric(A_m_n))
diagonal <- sqrt(eigen_B_n_n$values)
diagonal
```

```{r}
D_m_n <- diag(diagonal, nrow = m, ncol = n)
D_m_n
```

```{r}
U_m_m <- eigen_C_m_m$vectors
U_m_m
```

```{r}
V_n_n <- eigen_B_n_n$vectors
V_n_n
```

```{r}
U_m_m %*% D_m_n %*% t(V_n_n)
```

```{r}
R_svd <- svd(A_m_n, nv = ncol(A_m_n), nu = nrow(A_m_n)) 
R_svd
```

```{r}
R_svd$u %*% diag(R_svd$d, nrow = m, ncol = n) %*% t(R_svd$v)
```


## Reduced SVM
$$
A_{m \times n} = U_{m \times m} D_{m \times n} V_{n \times n}^T = 
U_{m \times r} D_{r \times r} V_{r \times n}^T
$$


```{r}
R_svd <- svd(A_m_n) 
R_svd
```

```{r}
R_svd$u %*% diag(R_svd$d) %*% t(R_svd$v)
```

## SVD based aproximation

$$
A_{m \times n} = U_{m \times m} D_{m \times n} V_{n \times n}^T = 
U_{m \times r} D_{r \times r} V_{r \times n}^T \approx \\
\approx U_{m \times k} D_{k \times k} V_{k \times n}^T
$$
where $k< r$

```{r}
k <- n - 1
R_svd$u[1:m, 1:k] %*% diag(R_svd$d)[1:k, 1:k] %*% t(R_svd$v)[1:k, 1:n]
```

